Supplement for: Carbon cycling in mature and regrowth forests globally: a macroecological synthesis based on the global Forest Carbon (ForC) database

### 2021-01-22 manuscript/ERL_global_C_review.Rmd
incorporate @bpbond 's comments

*Review Results/ Synthesis.*  *ForC v3.0* yielded a comprehensive picture of C cycling in the world's major forest biomes. The rate of C cycling generally increased from boreal to tropical regions in both mature and regrowth forests, whereas C stocks showed less directional variation. The majority of flux variables, together with most live biomass pools, increased significantly with stand age. There was generally good closure of C budgets for mature forests, whereas age trends and C budget closure in young forests remain less clearly resolved.

*Discussion.* As climate change accelerates, understanding and managing the carbon dynamics of forests is critical to forecasting, mitigation, and adaptation. This synthetic global overview of C stocks and fluxes across biomes and stand ages will help to advance these efforts.

@@

A robust understanding of forest impacts on global C cycling is essential. 
Annual gross CO~2~ sequestration in forests (gross primary productivity, $GPP$) is estimated at >69 Gt C yr^-1^ [@badgley_terrestrial_2019], or >7 times average annual fossil fuel emissions from 2009-2018 [9.5 $\pm$ 0.5 Gt C yr^-1^; @friedlingstein_global_2019].
Most of this enormous C sequestration is counterbalanced by CO~2~ releases to the atmosphere through ecosystem respiration ($R_{eco}$) or fire, with forests globally dominant as sources of both soil respiration [@warner_spatial_2019] and fire emissions [@van_der_werf_global_2017].
In recent years, the remaining CO~2~ sink averaged 3.2 $\pm$ 0.6 GtC yr^-1^ from 2009-2018, offsetting 29% of anthropogenic fossil fuel emissions [@friedlingstein_global_2019].
However, deforestation, estimated at ~1 Gt C yr^-1^ in recent decades [@pan_large_2011; @tubiello_carbon_2020], reduces the net forest sink to ~1.1-2.2 Gt C yr^-1^ across Earth's forests [@friedlingstein_global_2019].

@@

For more than half a century, researchers have sought to understand how forest carbon cycling varies across stands, including those of different biomes [e.g., @lieth_primary_1973; @luyssaert_co2_2007] and stand ages [e.g., @odum_strategy_1969; @luyssaert_old-growth_2008].
Over this time, an increasingly refined conceptual understanding of the elements of ecosystem C cycles has developed, as a growing number of variables have been defined [e.g., @chapin_reconciling_2006], along with appropriate measurement methods [e.g., @clark_measuring_2001]. 
New technology has also enabled researches to directly measure an expanding set of variables, notably including the development of continuous measurements of soil CO~2~ efflux (**REF-BEN**) and ecosystem-atmosphere CO~2~ exchange (**REFS**).
Standardization of estimation techniques has also increased; for example, there has been increasing standardization of the biomass allometries strongly influence estimates of any C cycle variable including woody biomass or changes therein [e.g., @chojnacky_updated_2014; @chave_improved_2014].
Further standardization has been made possible through research networks such as ForestGEO [@anderson-teixeira_ctfs-forestgeo_2015; @davies_forestgeo_2021], NEON [@schimel_neon_2007], and FLUXNET [@baldocchi_fluxnet_2001; @novick_ameriflux_2018].
Remote sensing technology has become increasingly useful for global- or regional-scale estimates of a few critical variables [e.g., aboveground biomass, $B_{ag}$: @saatchi_benchmark_2011; @hu_mapping_2016; @spawn_harmonized_2020; gross primary productivity, $GPP$: @li_mapping_2019], yet measurement and validation of most forest C stocks and fluxes necessarily requires intensive on-the-ground data collection.

@@

Forest C cycling varies enormously across biomes, which are artificial categorical distinctions convenient for categorizing the world's forests according to major differences in climate, vegetation, *etc.* 
Since the early 19th century, it has been recognized that climate plays a dominant role in shaping differences among forests on a global scale [@humboldt_essay_1807; @holdridge_determination_1947].
Global scale data synthesis has revealed that C fluxes including $GPP$, net primary productivity ($NPP$), and soil respiration ($R_{soil}$) decrease with latitude or, correspondingly, increase with mean annual temperature and, to a lesser extent, precipitation [Fig. `r schematic_fluxes_biome` **REFS**; @lieth_primary_1973; @luyssaert_co2_2007; @banbury_morgan_global_nodate].
C stocks of mature forests show less directional variation (Fig. `r schematic_fluxes_biome`).
On average, $B_{ag}$ tends to decrease with latitude, but not as dramatically as fluxes, and with the highest $B_{ag}$ forests in relatively cool, moist temperate regions [**REFS**; @keith_re-evaluation_2009]. 
In contrast, standing and downed dead wood ($DW_{standing}$ and $DW_{down}$, respectively) and the organic layer ($OL$) tend to accumulate more in colder climates where decomposition is slow relative to $NPP$ [**REFS**].

Correlative analyses relating C cycle variables to climate and other environmental variables have recently been taken to a new level through use of machine-learning algorithms that relate ground-based C cycle data to *global data on climate/soils/satellite data*, making it possible to create fine-scale global maps of C cycling [e.g., **REFS**; @cook-patton_mapping_2020].
*This approach can be particularly effective when paired with satellite data ...  (e.g., aboveground biomass: Saatchi et al 2011, Hu et al 2016, Spawn et al 2020, gross primary productivity, GPP: Li and Xiao 2019, ).* 
Yet, any such analysis is constrained by the quality and coverage of ground-based estimates of forest C fluxes or stocks. 
While estimates of some variables (e.g., $B_{ag}$, $GPP$, $NPP$, $R_{soil}$) are widely available, many remain poorly characterized (e.g., $DW$; autotrophic respiration, $R_{auto}$) --even at the coarse resolution of biomes.
This is a critical limitation not only for understanding forest C cycling, but also for quantifying climate change mitigation across forest biomes or ecozones [e.g., **REFS**].

@@

Stand age is another important axis of variation in forest C cycling (Fig. `r schematic_fs_age`).
In 1969, E.P. Odum's "The Strategy of Ecosystem Development" laid out predictions as to how forest energy flows and organic matter stocks vary with stand age  [@odum_strategy_1969].
Although the conceptualization of the C cycle in this paper is simplistic by current standards, the paper was foundational in framing the theory around which research on the subject still revolves [@corman_foundations_2019], and the basic framework still holds, *albiet* with modest modifications [@andersonteixeira_altered_2013].
Following stand-clearing disturbance, $GPP$, $NPP$, and biomass of leaves ($B_{foliage}$) and fine roots ($B_{root-fine}$) increase rapidly and thereafter remain relatively stable ($B_{foliage}$,$B_{root-fine}$, sometimes $GPP$) or decline slightly [$NPP$, sometimes $GPP$; e.g., *Goulden et al. 2011, ....  refs in * @andersonteixeira_altered_2013].
The decline in $NPP$ occurs because $R_{auto}$ increases relative to $GPP$ as forests age, corresponding to declining carbon use efficiency with stand age [@delucia_forest_2007; @collalti_forest_2020].
Heterotrophic respiration, most of which originates from the soil ($R_{het-soil}$) remains relatively constant with stand age *[Law et al., 2003; Pregitzer & Euskirchen, 2004; Goulden et al., 2011]*, with the result that net ecosystem production ($NEP=GPP-R_{eco}$, where $R_{eco}$ is total ecosystem respiration) is initially negative, increases to a maximum at intermediate ages, and declines thereafter *[Law et al., 2003; Pregitzer & Euskirchen, 2004; Zhou et al., 2006; Baldocchi, 2008; Luyssaert et al., 2008; Amiro et al., 2010; Goulden et al., 2011]*.
The result is that biomass accumulates rapidly in young forests, followed by a slow decline to near zero in old forests *[Lichstein et al., 2009; Yang et al., 2011;
Hember et al., 2012]*.
While these trends have been subject of fairly recent review [@andersonteixeira_altered_2013], *there is need for a synthetic, quantitaitive review.*

@@

In the past *couple of/several* decades, researchers have started asking how age trends--mostly in $B_{ag}$ or total biomass ($B_{tot}$) accumulation-- vary across biomes.
Early research on this theme showed that biomass accumulation rates during secondary succession increase with temperature on a global scale [**REFS**; @anderson_temperature-dependence_2006] and *with precipitation in the neotropics* [**REFS**; @chazdon_carbon_2016]. 
Most recently, @cook-patton_mapping_2020 reinforced these earlier findings with a much larger dataset and crated a high-resolution global map of estimated potential C accumulation rates.
However, there has been little synthesis of cross-biome differences in variables other than biomass and its accumulation rate (but see @cook-patton_mapping_2020 for $DW$, $OL$, and soil C accumulation in young stands).
Given the important role of secondary forests in the current and future global C cycle, concrete understanding of age trends in C fluxes and stocks and how these vary across biomes is critical to better understanding of the global C cycle.
Accurate estimates of C sequestration rates by regrowth forests are also critical for national greenhouse gas accounting under the IPCC framework [**REFS**] and to quantifying the value of regrwoth forests for climate change mitigation [@anderson-teixeira_greenhouse_2011; @goldstein_protecting_2020].

>>>>

*Review Results/ Synthesis.*   The rate of C cycling generally increased from boreal to tropical regions in both mature and regrowth forests, whereas C stocks showed less directional variation. The majority of flux variables, together with most live biomass pools, increased significantly with stand age. There was generally good closure of C budgets for mature forests, whereas age trends and C budget closure in young forests remain less clearly resolved.

*Discussion.* As climate change accelerates, understanding and managing the carbon dynamics of forests is critical to forecasting, mitigation, and adaptation. This comprehensive and synthetic global overview of C stocks and fluxes across biomes and stand ages will help to advance these efforts.

@@

A robust understanding of forest impacts on global C cycling is essential. 
Annual gross sequestration in forests (gross primary productivity, $GPP$) is estimated at >69 Gt C yr^-1^ [@badgley_terrestrial_2019], or >7 times average annual fossil fuel emissions from 2009-2018 [9.5 $\pm$ 0.5 Gt C yr^-1^; @friedlingstein_global_2019].
Most of this enormous C sequestration is counterbalanced by releases to the atmosphere through ecosystem respiration ($R_{eco}$) or fire, with forests globally dominant as sources of both soil respiration [@warner_spatial_2019] and fire emissions [@van_der_werf_global_2017].
In recent years, the remaining C sink averaged 3.2 $\pm$ 0.6 Gt C yr^-1^ from 2009-2018, offsetting 29% of anthropogenic fossil fuel emissions [@friedlingstein_global_2019].
However, deforestation, estimated at ~1 Gt C yr^-1^ in recent decades [@pan_large_2011; @tubiello_carbon_2020], reduces the net forest sink to ~1.1-2.2 Gt C yr^-1^ across Earth's forests [@friedlingstein_global_2019].

@@

For more than half a century, researchers have sought to understand how forest carbon cycling varies across stands, including those of different biomes [e.g., @lieth_primary_1973; @luyssaert_co2_2007] and stand ages [e.g., @odum_strategy_1969; @luyssaert_old-growth_2008].
Over this time, an increasingly refined conceptual understanding of the elements of ecosystem C cycles has developed, as a growing number of variables have been defined [e.g., @chapin_reconciling_2006], along with appropriate measurement methods [e.g., @clark_measuring_2001]. 
New technology has also enabled researchers to directly measure an expanding set of variables, notably including the development of continuous measurements of soil CO~2~ efflux [@kuzyakov_sources_2006] and ecosystem-atmosphere CO~2~ exchange [@baldocchi_fluxnet_2001].
Measurement techniques have been increasingly standardized; for example, of the biomass allometries that strongly influence estimates of most C cycle variables [e.g., @chojnacky_updated_2014; @chave_improved_2014].
Further standardization has been made possible through research networks such as ForestGEO [@anderson-teixeira_ctfs-forestgeo_2015; @davies_forestgeo_2021], NEON [@schimel_neon_2007], and FLUXNET [@baldocchi_fluxnet_2001; @novick_ameriflux_2018].
Remote sensing technology has become increasingly useful for global- or regional-scale estimates of a few critical variables [e.g., aboveground biomass, $B_{ag}$: @saatchi_benchmark_2011; @hu_mapping_2016; @spawn_harmonized_2020; gross primary productivity, $GPP$: @li_mapping_2019], yet measurement and validation of most forest C stocks and fluxes necessarily requires intensive on-the-ground data collection.

@@

Forest C cycling varies enormously across biomes, which are artificial categorical distinctions convenient for categorizing the world's forests according to major differences in climate, vegetation, *etc.* 
Since the early 19th century, it has been recognized that climate plays a dominant role in shaping differences among forests on a global scale [@humboldt_essay_1807; @holdridge_determination_1947].
Global scale data syntheses have shown that C fluxes including $GPP$, net primary productivity ($NPP$), and soil respiration ($R_{soil}$) decrease with latitude or, correspondingly, increase with mean annual temperature and, to a lesser extent, precipitation [Fig. `r schematic_fluxes_biome` **REFS**; @lieth_primary_1973;  @luyssaert_co2_2007; @hursh_sensitivity_2017; @banbury_morgan_global_nodate].
C stocks of mature forests show less directional variation (Fig. `r schematic_fluxes_biome`).
On average, $B_{ag}$ tends to decrease with latitude, but not as dramatically as fluxes, and with the highest $B_{ag}$ forests in relatively cool, moist temperate regions [**REFS**; @keith_re-evaluation_2009]. 
In contrast, standing and downed dead wood ($DW_{standing}$ and $DW_{down}$, respectively) and the organic layer ($OL$) tend to accumulate more in colder climates where decomposition is slow relative to $NPP$ [@harmon_ecology_1986; @allen_global_2002].

Correlative analyses relating C cycle variables to climate and other environmental variables have recently been taken to a new level through use of machine-learning algorithms that relate ground-based C cycle data to *global data on climate/soils/satellite data*, making it possible to create fine-scale global maps of C cycling [e.g., **REFS**;@warner_spatial_2019; @cook-patton_mapping_2020].
*This approach can be particularly effective when paired with satellite data ...  (e.g., aboveground biomass: Saatchi et al 2011, Hu et al 2016, Spawn et al 2020, gross primary productivity, GPP: Li and Xiao 2019, ).* 
Any such analysis is however constrained by the quality and coverage of ground-based estimates of forest C fluxes or stocks. 
While estimates of some variables (e.g., $B_{ag}$, $GPP$, $NPP$, $R_{soil}$) are widely available, many remain poorly characterized (e.g., $DW$; autotrophic respiration, $R_{auto}$) --even at the coarse resolution of biomes.
This is a critical limitation not only for understanding forest C cycling, but also for quantifying climate change mitigation across forest biomes or ecozones [e.g., **REFS**].

@@

Stand age is another important axis of variation in forest C cycling (Fig. `r schematic_fs_age`).
In 1969, E.P. Odum's "The Strategy of Ecosystem Development" laid out predictions as to how forest energy flows and organic matter stocks vary with stand age  [@odum_strategy_1969].
Although the conceptualization of the C cycle in this paper is simplistic by current standards, the paper was foundational in framing the theory around which research on the subject still revolves [@corman_foundations_2019], and the basic framework still holds, albeit with modest modifications [@andersonteixeira_altered_2013].
Following stand-clearing disturbance, $GPP$, $NPP$, and biomass of leaves ($B_{foliage}$) and fine roots ($B_{root-fine}$) increase rapidly and thereafter remain relatively stable ($B_{foliage}$,$B_{root-fine}$, sometimes $GPP$) or decline slightly [$NPP$, sometimes $GPP$; e.g., *Goulden et al. 2011, ....  refs in * @andersonteixeira_altered_2013].
The decline in $NPP$ occurs because $R_{auto}$ increases relative to $GPP$ as forests age, corresponding to declining carbon use efficiency with stand age [@delucia_forest_2007; @collalti_forest_2020].
Heterotrophic respiration, most of which originates from the soil ($R_{het-soil}$) remains relatively constant with stand age *[Law et al., 2003; Pregitzer & Euskirchen, 2004; Goulden et al., 2011]*, with the result that net ecosystem production ($NEP=GPP-R_{eco}$, where $R_{eco}$ is total ecosystem respiration) is initially negative, increases to a maximum at intermediate ages, and declines thereafter *[Law et al., 2003; Pregitzer & Euskirchen, 2004; Zhou et al., 2006; Baldocchi, 2008; Luyssaert et al., 2008; Amiro et al., 2010; Goulden et al., 2011]*.
The result is that biomass accumulates rapidly in young forests, followed by a slow decline to near zero in old forests *[Lichstein et al., 2009; Yang et al., 2011;
Hember et al., 2012]*.
While these trends have been subject of fairly recent review [@andersonteixeira_altered_2013], *there is need for a synthetic, quantitaitive review.*

@@

In the past several decades, researchers have started asking how age trends--mostly in $B_{ag}$ or total biomass ($B_{tot}$) accumulation-- vary across biomes.
Early research on this theme showed that biomass accumulation rates during secondary succession increase with temperature on a global scale [**REFS**; @anderson_temperature-dependence_2006] and *with precipitation in the neotropics* [**REFS**; @chazdon_carbon_2016]. 
Most recently, @cook-patton_mapping_2020 reinforced these earlier findings with a much larger dataset and crated a high-resolution global map of estimated potential C accumulation rates.
However, there has been little synthesis of cross-biome differences in variables other than biomass and its accumulation rate (but see @cook-patton_mapping_2020 for $DW$, $OL$, and soil C accumulation in young stands).
Given the important role of secondary forests in the current and future global C cycle, concrete understanding of age trends in C fluxes and stocks and how these vary across biomes is critical to better understanding of the global C cycle.
Accurate estimates of C sequestration rates by regrowth forests are also critical for national greenhouse gas accounting under the IPCC framework [**REFS**] and to quantifying the value of regrwoth forests for climate change mitigation [@anderson-teixeira_greenhouse_2011; @goldstein_protecting_2020].
